---
title: 'Predicting Flake Mass: A View from Machine Learning. Lithic Technology'
author: "Guillermo Bustos-Pérez"
date: "7/3/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1) Load packages, read and check data    

The data is available in this repository as a **.csv** file. Please note that original data uses **","** as decimal marker instead of using **"."**. Thus, it is required to use the function **read_csv2**.  

```{r}
# Load packages  
library(tidyverse)
```
```{r message=FALSE, warning=FALSE, include=FALSE}
library(kableExtra)
```

```{r}
# Read in data
Reg_Data <- read.csv2("Data.csv")
```

Please note that here the function **kable()** from package **kableExtra** is being employed to visualize imported data. Function **head()** can also be employed.

```{r echo=FALSE}
# Instead of head
kable(Reg_Data[1:5,],
      col.names = gsub("[_]", " ", names(Reg_Data)))
```
```{r}
# Get column names
colnames(Reg_Data)
```
## Descriptive statistics of experimental assemblage   

The sample consisted of 300 freehand experimental flint flakes knapped using a hard hammer. Flakes belonged to nearly 20 knapping sequences wherein a wide variety of knapping methods were employed – hierarchical (Levallois and hierarchical discoid), bifacial (discoid), and unipolar – to generate the experimental sample, ensuring a wide range of morphologies. All selected flakes from the knapping sequences were complete, and all presented feather terminations. Since a key aspect of the experimentation was to estimate flake mass and independently of exterior factors, hammerstones included a wide selection of limestone, sandstone, and quartzite pebbles, allowing for a diverse range of morphologies and potential active percussion areas.

Summary statistics of the experimental assemblage. Note that **"Platform size1"** refers to platform size measured following Muller and Clarkson (2016); while **"Platform size2"** refers to measures following Andrefsky (2005).

```{r}
# Make data frame of descriptive statistics of assemblage
Summary_Assem <- data.frame(
    rbind(
      data.frame(data.matrix(summary(Reg_Data$Length))) %>% t(),
      data.frame(data.matrix(summary(Reg_Data$Width))) %>% t(),
      data.frame(data.matrix(summary(Reg_Data$Mean_Thick))) %>% t(),
      data.frame(data.matrix(summary(Reg_Data$Surf_Plat))) %>% t(),
      data.frame(data.matrix(summary(Reg_Data$Surf_Plat_II))) %>% t(),
      data.frame(data.matrix(summary(Reg_Data$Weight))) %>% t()
      ))

# Place column names for each measure
Measure <- c("Length", "Width", "Mean Thickness", "Platform Surface1",
             "Platform Surface2", "Weight")

Summary_Assem <- cbind(Measure, Summary_Assem)

rownames(Summary_Assem) <- 1:nrow(Summary_Assem)
```
```{r echo=FALSE}
# Instead of head
kable(Summary_Assem)
```


```{r, eval=FALSE}
Reg_Data_2 <- Reg_Data %>% 
  mutate(Log_Weight = log10(Weight),
         Log_Thick = log10(MeanThick),
         Log_SD_Thick = log10(SDThick),
         Log_Plat = log10(Surface.Plat),
         Log_Plat_2 = log10(Surf_Plat_II),
         Log_Plat_De = log10(`Platform Deepnes`),
         Log_EPA = log10(EPA),
         Log_IPA = log10(IPA)) 
```


## References   


